The effect of starvation on the dynamics of consumer populations

### 2016-06-15 Manuscript/Yeakel_Kempes_Redner-5-16-2016.tex
New analyses with allometric rates

\title{The effect of starvation on the dynamics of consumer populations}
\author{Yeakel Kempes Redner}
\maketitle

@@

The behavioral ecology of most, if not all, organisms is influenced by the energetic state of individuals.
An individual's energetic state directly influences how it invests its stores in an uncertain environment.
Such behaviors are generally manifested as trade-offs, which often concern investing in individual maintenance and growth vs. producing offspring [Mangel], among a host of other behavioral duties [REFS].
The timing of these behaviors is often important and is under strong selective pressure, as they tend to have large effects on the future fitness of the organism [Mangel].
%For example, rotifers that reproduce have lower survival rates than those that don't when they are under nutritional stress.
To what extent, and when, organisms invest in these two necessary biological functions -- growth and maintenance vs. reproduction -- may be driven by habitat, seasonality, evolutionary history, inter- or intra-specific interactions, or even resource limitation.
The influence of resource limitation on an organism's ability to maintain its nutritional stores may lead to delays or shifts in reproduction. %and the subsequent loss of nutritional reserves
%Such tradeoffs are also bound to have population-level consequences, although this has received little theoretical or empirical attention.

%The reproductive vs. storage tradeoff
Maximizing fitness between growth and maintenance activities vs. reproductive behaviors in large part structures the life-history of species, and this can be achieved by a variety of potential mechanisms all of which, to some extent, depend on resource availability.
%Reproduction can be separated from somatic growth/maintenance by a variety of potential mechanisms:
%Multiple mechanisms contribute to the differentiation of these behaviors:
Behavioral: The investment of time and energy towards reproductive and parental behaviors depends on resource availability \citep{Morris:1987eo}.
%, while stress can generally lead to reprodusuppression of reproductive behaviors \citep{Morgan:1999do}. 
For example, reindeer invest less in calves born after harsh winters (when the mother's energetic state is poor) than in calves born after moderate winters \citep{Tveraa:2003fq}, whereas many bird species invest differently in broods during periods of resource scarcity \citep{Daan:1988va,Jacot:2009dv}, sometimes delaying or foregoing reproduction for a breeding season \citep{Barboza:2002in}.
Resource limitation can also alter the behaviors of species in well-mixed environments: freshwater and marine zooplankton have been observed to avoid reproduction under nutritional stress \citep{Threlkeld:1976ih}, while those that do reproduce evince lower survival rates \citep{Kirk:1997cc}.
Similarly, artificially induced stress has been observed to decrease reproductive success in Atlantic cod \citep{Morgan:1999do}.
Physiological: Diverse mammals (47 species in 10 families) exhibit delayed implantation whereby females postpone fetal development (blastocyst implantation), often timing with accumulation of nutritional reserves \citep{Mead:1989dt,Sandell:1990kw}. %, including multiple species of Grizzly bears (\emph{Ursus horribilis}) and the European badger (\emph{Meles melesc}) \citep{Woodroffe:1995id
Furthermore, many mammals (including humans) suffer irregular menstrual cycling and higher rates of spontaneous abortion during periods of nutritional stress \citep{Bulik:1999eo,Trites:2003cc}.
Spatio-temporal: Organisms may also separate maintenance/growth from reproduction over space and time. 
For example, many salmonids, birds, and some mammals return to migratory breeding sites to reproduce after one or multiple seasons in alternative environments spent accumulating body mass and nutritional reserves \citep{Weber:1998jg,Mduma:1999cp,Moore:2014hi}.
%great tits invest more energy into broods when resource abundance is greater, while food availability is generally seen to correlate with clutch size among many bird species \citep{Daan:1988va}.
%The mechanisms that control reproductive effort as a function of energetic reserves differ among species.
%Even humans alter their growth trajectories and have higher chances of fetal abortion during periods of starvation.
%Thus, some organisms partition reproductive vs. foraging effort over both space and time, whereas other rely on physiological mechanisms to optimize reproductive success against survival when adequate resources are unavailable.
The existence of so many independently evolved mechanisms across such a diverse suite of organisms points to the importance and universality of the fundamental tradeoff between spending energy on tissues vs. spending energy on passing down genetic material.

%To population-level effects
Organisms employ different strategies to avoid reproduction during times of nutritional stress, and how this is achieved has received tremendous empirical and theoretical attention, owing to the importance of these activities in shaping life-history [REF].
Less well understood is how resource limitation and these behavioral/physiological tradeoffs affect dynamics at the level of the population.
% when organisms are able to avoid the risk of spending energy on reproduction when starvation is likely.
Traditional Lotka-Volterra models assume a dependence of consumer population growth rates on resource density, thus implicitly incorporating the requirement of resource availability for reproduction.
Although this implicit dependence connects resource limitation to lower consumer growth rates, the following biological realities are not taken into account:
\emph{i}) some individuals experience nutritional stress at a given time and under a given set of external conditions, while others do not; those that do have multiple pathways enabling reproductive cessation, 
\emph{ii}) the portion of the population that is not nutritionally stressed reproduces at a near-constant rate, constrained by allometric relationships, and
\emph{iii}) the rate at which individuals transition from nutritionally poor to replete states (and vice versa) are governed by the organism's metabolism, which is also strongly constrained by allometry.
To incorporate these important elements, resource limitation and the subsequent effects of starvation can be accounted for \emph{explicitly}, such that reproductive growth of a population is only allowed to occur for individuals with sufficient energetic reserves.
Incorporating energetic dynamics that occur at an individual level [REF] into a population-based framework \citep{Kooi2000,Sousa:2010ez}, though straightforward conceptually, has many challenges that arise mathematically \citep{Diekmann:2010da}, and this has limited simple theoretical models aiding our understanding of larger-scale dynamics.

%What we did...
Here we explore how reproductive trade-offs, which occur at the level of the individual, may influence the dynamics of a population.
We first establish a simple stage-structured population model that captures the essential dynamics of energetic reproductive tradeoffs, and explore the impact of different fluxes on system-level stability.
By relating different rate constants to allometric relationships, we uncover important constraints in the timescales of different physiological processes that determine dynamics and investigate how organisms of alternative body sizes and taxonomic affinities are expected to evince contrasting population-level dynamics.
We then develop a more general dynamic model in order to understand if there are common attributes of these energetically-constrained systems that have implications beyond assumptions inherent to the specific model.
We show [WHAT]?
%Finally, we compare our population-dynamic approach with a stochastic dynamic program that incorporates different individual-level behaviors as a function of energetic state.

@@

\begin{align}
\frac{\rm d}{\rm dt} F &= \lambda F + \rho HR - \sigma (1-R)F, \nonumber \\
\frac{\rm d}{\rm dt} H &= \sigma (1-R)F - \rho RH - \mu H, \nonumber \\
\frac{\rm d}{\rm dt} R &= a R(1-R) - R(\rho H+ b F).
\label{eq_specific}
\end{align}

@@

\begin{align}
F^* &= \frac{a  \lambda  \mu  (\mu +\rho )}{(\lambda  \rho +\mu  \sigma ) (\lambda  \rho +\mu  m)}, \nonumber \\
H^* &= \frac{a  \lambda ^2 (\mu +\rho )}{(\lambda  \rho +\mu  \sigma ) (\lambda  \rho +\mu  m)}, \nonumber \\
R^* &= \frac{\mu  (\sigma -\lambda )}{\lambda  \rho +\mu  \sigma }.
\end{align}

@@

%Transcritical bifurcation
If the parameters of the Jacobian matrix at the internal steady state are such that its leading eigenvalue is $<0$, then the system is stable to small pulse perturbations, conditioned on the value of the starvation rate $\sigma$ relative to the value of the consumer reproduction rate $\lambda$.
As $\sigma$ nears and becomes lower than a given $\lambda$, the resource steady state $R^*$ crosses the origin and exchanges stability to become unstable.
As such, a transcritical bifurcation exists at $\lambda = \sigma$, such that the existence of an internal stable fixed point is dependent on the condition that $\sigma > \lambda$.
Biologically, this means that the rate of starvation is greater (operating on a smaller timescale) than the rate of consumer reproduction (operating on a relatively longer timescale).
As will be shown in Section XX, this general expectation will hold for most classes of organisms, while the exact difference in timescales between reproduction and starvation can be derived using allometric scaling relationships.
%Kempes (REF) has shown that population growth rates scale to $M^{\alpha-1}$ (where $M$ is the reproductive mass of an organism and $\alpha$ is the scaling exponent), whereas starvation is necessarily bound to shorter timescales, which will be elaborated on below.

%Hopf bifurcation
Oscillating, or cyclic, dynamics present additional risks to populations.
If cycles are large, stochastic effects may result in extinction.
In continuous-time systems, cycles arise when a pair of complex conjugate eigenvalues cross the imaginary axis and attain positive real parts.
%Of additional interest is the existence of parameter regions that permit the existence of oscillatory, or cyclic, dynamics.
This condition is called a Hopf bifurcation, and is defined by ${\rm Det}({\bf S}) = 0$, where $\bf S$ is the Sylvester matrix, which is composed of the coefficients of the characteristic polynomial describing the Jacobian matrix.
Although the Hopf condition for the specific 2-stage model cannot be easily written, the analytical solution can be explored using a symbolic computational language such as \emph{Mathematica}.

@@

Nearly all of the rates described in the specific model, and generalized upon in the generalized model, are to some extent governed by the body size of the consumer.
The scaling relationship between an organism's metabolic rate $B$ and its body size at reproductive maturity $M$ plays a central role in other scaling relationships.
Organismal metabolic rate $B$ is known to scale as $B = B_0 M^\eta$, where $\eta$ is the scaling exponent, generally assumed to be 3/4 for metazoans, etc.
Kempes et al. [REF] show how the population-level growth rate also can be related to body size as $\lambda = \lambda_0 M^{\eta-1}$.

Population growth requires that individuals

@@

%Hopf bifurcation and oscillations sigma vs. rho
The 2-stage consumer resource model exhibits two qualitatively different behavioral regimes.
Because portions of the consumer population exist in either full or hungry states simultaneously, the internal fixed point can either be a stable equilibrium, or exhibit sustained oscillatory behavior, depending on the rates of starvation $\sigma$ and recovery $\rho$.
The transition from stable non-oscillatory dynamics to oscillatory dynamics occurs at the Hopf bifurcation condition where two complex conjugate eigenvalues cross the imaginary axis and attain positive real parts.
Although there is an analytical solution to the Hopf bifurcation condition, it cannot be written efficiently
When the starvation rate is low, oscillatory dynamics are more likely to occur for a given value of the recovery rate.
This can be understood intuitively: for low starvation rates, resources are depressed by an infusion of full consumers, which subsequently starve thereby allowing the resource to recover and continuing the cycle.
When the starvation rate is high, the response of consumer growth to resource abundance is muted, such that oscillations tend to decay over time.
Thus, higher starvation rates $\sigma$ desensitizes changes in the consumer population to changes in resource density, and lower rates of recovery $\rho$ amplifies this effect.

@@

%The reproductive tradeoff as a strategy
%

@@

 \begin{figure}[h]
 	\centering
 	\includegraphics[width=0.5\textwidth]{fig_HopfPlotb.pdf}
 	\caption{
 	}
 	\label{Hopfb}
 \end{figure}
%
%
% \begin{figure}[h]
% 	\centering
% 	\includegraphics[width=0.5\textwidth]{fig_resvuln.pdf}
% 	\caption{
% 	Probability that the resource value is less than threshold = 0.05
% 	}
% 	\label{SN}
% \end{figure}
%
%
%
% \begin{figure}[h]
% 	\centering
% 	\includegraphics[width=0.5\textwidth]{fig_Competition.pdf}
% 	\caption{
% 	Hopf bifurcation
% 	}
% 	\label{SN}
% \end{figure}

>>>>

\title{The effect of starvation on the dynamics of consumer populations}
\author{Yeakel, Kempes, \& Redner}
\maketitle

@@

The behavioral ecology of most, if not all, organisms is influenced by the energetic state of individuals.
An individual's energetic state directly influences how it invests its stores in an uncertain environment.
Such behaviors are generally manifested as trade-offs, which often concern investing in individual maintenance and growth or allocating energy towards reproduction \citep{Kirk:1997cc,Kempes:2012hy}. %among a host of other behavioral duties [REFS].
The timing of these behaviors is often important and is under strong selective pressure, as they tend to have large effects on the future fitness of the organism \citep{Mangel:1988uaa}.
%For example, rotifers that reproduce have lower survival rates than those that don't when they are under nutritional stress.
To what extent, and when, organisms invest in these two necessary biological functions -- growth and maintenance vs. reproduction -- may be driven by habitat, seasonality, evolutionary history, inter- or intra-specific interactions, as well as resource limitation.
Importantly, the influence of resource limitation on an organism's ability to maintain its nutritional stores may lead to repeated delays or shifts in reproduction over the course of an organism's life. %and the subsequent loss of nutritional reserves
%Such tradeoffs are also bound to have population-level consequences, although this has received little theoretical or empirical attention.

%The reproductive vs. storage tradeoff
Maximizing fitness between growth and maintenance activities vs. reproductive behaviors in large part structures the life-history of species, and this can be achieved by a variety of potential mechanisms, all of which, to some extent, depend on resource availability.
[In extreme case of unicellular organisms, nutrition is unavoidably linked to reproduction because the nutritional state of the individual regulates all aspects of the cell cycle.]
%Reproduction can be separated from somatic growth/maintenance by a variety of potential mechanisms:
%Multiple mechanisms contribute to the differentiation of these behaviors:
Behavioral: The investment of time and energy towards reproductive and parental behaviors depends on resource availability \citep{Morris:1987eo}.
%, while stress can generally lead to reprodusuppression of reproductive behaviors \citep{Morgan:1999do}. 
For example, reindeer invest less in calves born after harsh winters (when the mother's energetic state is poor) than in calves born after moderate winters \citep{Tveraa:2003fq}, whereas many bird species invest differently in broods during periods of resource scarcity \citep{Daan:1988va,Jacot:2009dv}, sometimes delaying or foregoing reproduction for a breeding season \citep{Barboza:2002in}.
%Resource limitation can also alter the behaviors of species in well-mixed environments: 
Freshwater and marine zooplankton have been observed to avoid reproduction under nutritional stress \citep{Threlkeld:1976ih}, with those that do reproduce evincing lower survival rates \citep{Kirk:1997cc}, while artificially induced stress has been observed to decrease reproductive success in Atlantic cod \citep{Morgan:1999do}.
Physiological: Diverse mammals (47 species in 10 families) exhibit delayed implantation whereby females postpone fetal development (blastocyst implantation) to time with accumulation of nutritional reserves \citep{Mead:1989dt,Sandell:1990kw}. %, including multiple species of Grizzly bears (\emph{Ursus horribilis}) and the European badger (\emph{Meles melesc}) \citep{Woodroffe:1995id
Furthermore, many mammals (including humans) suffer irregular menstrual cycling and higher rates of spontaneous abortion during periods of nutritional stress \citep{Bulik:1999eo,Trites:2003cc}.
Spatio-temporal: Organisms may also separate maintenance/growth from reproduction over space and time. 
For example, many salmonids, birds, and some mammals return to migratory breeding sites to reproduce after one or multiple seasons in alternative environments spent accumulating body mass and nutritional reserves \citep{Weber:1998jg,Mduma:1999cp,Moore:2014hi}.
%great tits invest more energy into broods when resource abundance is greater, while food availability is generally seen to correlate with clutch size among many bird species \citep{Daan:1988va}.
%The mechanisms that control reproductive effort as a function of energetic reserves differ among species.
%Even humans alter their growth trajectories and have higher chances of fetal abortion during periods of starvation.
%Thus, some organisms partition reproductive vs. foraging effort over both space and time, whereas other rely on physiological mechanisms to optimize reproductive success against survival when adequate resources are unavailable.
The existence of so many independently evolved mechanisms across such a diverse suite of organisms points to the importance and universality of the fundamental tradeoff between energetic expenditures on the growth and maintenance of tissues vs. reproductive behaviors.

%To population-level effects
The different strategies employed to avoid reproduction during times of nutritional stress, and how this is achieved among different organisms, has received tremendous empirical and theoretical attention owing to the importance of these activities in shaping life-history [REF].
Less well understood is how resource limitation and these behavioral/physiological tradeoffs affect dynamics at the level of the population.
% when organisms are able to avoid the risk of spending energy on reproduction when starvation is likely.
Traditional Lotka-Volterra models assume a dependence of consumer population growth rates on resource density, thus \emph{implicitly} incorporating the requirement of resource availability for reproduction.
Although this implicit dependence connects resource limitation to lower consumer growth rates, the following biological realities are not taken into account: 
\begin{itemize}
\item some individuals experience nutritional stress at a given time and under a given set of external conditions, while others do not; those that do have multiple pathways enabling reproductive cessation, 
\item the portion of the population that is not nutritionally stressed is expected to reproduce at a near-constant rate, and
\item the rates at which individuals transition from nutritionally poor to replete states and back are governed by the organism's metabolic capabilities of growing and maintaining tissue, which is allometrically constrained \citep{Kempes:2012hy}.
\end{itemize}
[The neglect of these features has important dynamic shortcomings:]
[Here we use body size and allometries to capture a wide variety of taxonomic diversity]
[There isn't a deep understanding with how models vary across different life histories, traits, etc... how do equations change with allometric constraints for different class of organisms? 
i.e. Use body size as a generating parameter to capture a wide variety of diversity.]

To incorporate these elements, resource limitation and the subsequent effects of starvation may be accounted for \emph{explicitly}, such that reproductive growth of a population is permitted only for individuals with sufficient energetic reserves.
Incorporating energetic dynamics that occur at an individual level \citep{Kooi2000} into a population-based framework \citep{Kooi2000,Sousa:2010ez}, though straightforward conceptually, quickly runs into mathematical obstacles \citep{Diekmann:2010da}.
[An alternative approach is to start with the macroscale patterns in order to understand certain tradeoffs etc]
The complexities associated with scaling individual-level energetics to that of a population has thus limited the development of theoretical models that may aid our understanding of the effects of such tradeoffs on population dynamics.

%What we did...
Here we explore how the energetic tradeoff between maintaining and building tissue and reproduction, which occurs at the level of the individual, may influence the dynamics of a population.
We first establish a simple nutritional state-structured population model that captures the essential dynamics of energetic reproductive tradeoffs, and explore the impact of different fluxes on system-level stability.
By relating different rate constants to allometric constraints, we uncover important relationships between the timescales of physiological and reproductive processes, and show how organisms of different body sizes and taxonomic affinities are expected may be prone to different dynamics.
The general rate laws that control both reproductive and energetic processes of the population, we show that

%We then develop a more general model to understand if there are common attributes of these energetically-constrained systems that have implications beyond assumptions inherent to the specific model.
We show [WHAT]?
%Finally, we compare our population-dynamic approach with a stochastic dynamic program that incorporates different individual-level behaviors as a function of energetic state.

@@

\begin{align}
\begin{split}
\frac{\rm d}{\rm dt} F &= \lambda F + \rho RH - \sigma (1-R)F,  \\
\frac{\rm d}{\rm dt} H &= \sigma (1-R)F - \rho RH - \mu H,  \\
\frac{\rm d}{\rm dt} R &= \alpha R(1-R) - R(\rho H+ \beta F).
\label{eq_specific}
\end{split}
\end{align}

[NSM = Nutritional State Model]
[LVM = Lotka Volterra Model]

@@

\begin{align}
\begin{split}
F^* &= \frac{a  \lambda  \mu  (\mu +\rho )}{(\lambda  \rho +\mu  \sigma ) (\lambda  \rho +\mu  m)}, \\
H^* &= \frac{a  \lambda ^2 (\mu +\rho )}{(\lambda  \rho +\mu  \sigma ) (\lambda  \rho +\mu  m)}, \\
R^* &= \frac{\mu  (\sigma -\lambda )}{\lambda  \rho +\mu  \sigma }.
\end{split}
\end{align}

@@

%Transcritical bifurcation
If the parameters of the Jacobian matrix at the internal steady state are such that its leading eigenvalue is $<0$, then the system is stable to small pulse perturbations, conditioned on the value of the starvation rate $\sigma$ relative to the value of the consumer reproduction rate $\lambda$.
As $\sigma$ nears and becomes lower than a given $\lambda$, the resource steady state $R^*$ crosses the origin and exchanges stability to become unstable.
[Explain what a transcritical bifurcation means in terms of the dynamics] From a regime where resources are negative and consumer populations become infinite to one that is biological feasable]
%As such, a transcritical bifurcation exists at $\lambda = \sigma$, such that the existence of an internal stable fixed point is dependent on the condition that $\sigma > \lambda$.
This means that the rate of starvation must be greater (operating on a smaller timescale) than the rate of consumer reproduction (operating on a relatively longer timescale).
This general expectation will hold for most classes of organisms, which we will show by deriving allometric scaling relationship that serve to constrain the covariation of parameter values.
% while the exact difference in timescales between reproduction and starvation can be derived using allometric scaling relationships.
%Kempes (REF) has shown that population growth rates scale to $M^{\alpha-1}$ (where $M$ is the reproductive mass of an organism and $\alpha$ is the scaling exponent), whereas starvation is necessarily bound to shorter timescales, which will be elaborated on below.

%Hopf bifurcation
[In addition to the hard bound defined by the TC bifurcation, oscillating or cyclic dynamics represent an implicit constraint by putting populations at risk... also present certain constraints to the feasibility of populations]
%Oscillating, or cyclic, dynamics present additional risks to populations.
If cycles are large, stochastic effects may result in extinction.
In continuous-time systems, cycles arise when a pair of complex conjugate eigenvalues cross the imaginary axis and attain positive real parts.
%Of additional interest is the existence of parameter regions that permit the existence of oscillatory, or cyclic, dynamics.
This condition is called a Hopf bifurcation, and is defined by ${\rm Det}({\bf S}) = 0$, where $\bf S$ is the Sylvester matrix, which is composed of the coefficients of the characteristic polynomial describing the Jacobian matrix.
%Although the Hopf condition for the specific 2-stage model cannot be easily written, the analytical solution can be explored using a symbolic computational language such as \emph{Mathematica}.

@@

[Link Allometry stuff to our model - what does it provide to the story]
[Introduce biological and linked constraints]
[Allometries capture vast amounts of diversity via a single parameter --- body size]
[Allometries have captured int terest etc across multiple scales and biological classes of organisms]

Nearly all of the rates described in the specific model, and later in the generalized model, are to some extent governed by the body size of the consumer. For example, the scaling relationship between an organism's metabolic rate $B$ and its body size at reproductive maturity $M$ is well documented \cite{} and plays a central role in a variety of scaling relationships.
Organismal metabolic rate $B$ is known to scale as $B = B_0 M^\eta$, where $\eta$ is the scaling exponent, generally assumed to be $3/4$ for metazoans, and varies in unicellular species between $\eta\approx 1$ in eukaryotes and $\eta\approx 1.76$ in bacteria \cite{delong}. Several efforts have shown how a partitioning of this metabolic rate between growth and maintenance purposes can be used to derive a general equation for the growth trajectories and growth rates of organisms ranging from bacteria to metazoans \cite{kempes}. More specifically, the interspecific trends in growth rate can be approximated by $\lambda = \lambda_0 M^{\eta-1}$. This relationship is derived from the simple balance
 \begin{equation}
 B_{0}m^{\alpha}=E_{m}\frac{dm}{dt}+B_{m}m
 \label{balance}
 \end{equation}
 [a and b notation --- these parameters are easily measured bioenergetic parameters which are often approximately invariant across organisms of vastly different size. Our notation seeks to illustrate that the allometric model fundamentally depends on a small number of free parameters.]
 where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is the metabolic rate to support an existing unit of mass, and $m$ is the mass at any point in development. It is useful to explicitly write this balance because it can also be modified to understand the rates of both starvation and recovery from starvation. [Spell out the connection to nutritional state more explicitly] [As we will see it is possible to derive both sigma and rho from this balance]

For the rate of starvation, we make the simple assumption that an organism must meet its maintenance requirements using digested mass as the sole energy source. This assumption implies the simple metabolic balance 
\begin{equation}
\frac{dm}{dt}E_{m}^{\prime}=-B_{m}m
\end{equation}
where $E_{m}^{\prime}$ is the amount of energy stored in a unit of existing body mass which may differ from  $E_{m}$, the energy required to synthesis a unit of biomass. Give the adult mass,  $M$, of an organism this energy balance prescribes the mass trajectory of a starving organism:
\begin{equation}
m\left(t\right)=Me^{-B_{m}t/E_{m}^{\prime}}.
\end{equation}
Considering that only certain tissues can be digested for energy, for example the brain cannot be degraded to fuel metabolism, we define the rate for starvation and death by the timescales required to reach specific fractions of normal adult mass. We define $m_{starve}=\epsilon M$ where it could be the case that organisms have a systematic size-dependent requirement for essential tissues, such as the minimal bone or brain mass. For example, considering the observation that body fat in mammals scales with overall body size according to $M_{f}=f_{0}M^{\gamma}$, and assuming that once this mass is fully digested the organism begins to starve, would imply that $\epsilon=1-f_{0}M^{\gamma}/M$. Taken together the time scale for starvation is given by
\begin{equation}
t_{\sigma}=-\frac{E_{m}\log\left(\epsilon\right)}{B_{m}}.
\end{equation}
The starvation rate is $\sigma=1/t_{s}$, which implies that $\sigma$ is independent of adult mass if $\epsilon$ is a constant, and if $\epsilon$ does scale with mass, then $\sigma$ will have a factor of $1/\log\left(1-f_{0}M^{\gamma}/M\right)$. In either case $\sigma$ does not have a simple scaling with $\lambda$ which is important for the dynamics that we later discuss.

The time to death should follow a similar relationship, but defined by a lower fraction of adult mass, $m_{death}=\epsilon^{\prime} M$. Consider, for example, that an organism dies once it has digested all fat and muscle tissues, and that muscle tissue scales with body mass according to $M_{mm}=mm_{0}M^{\zeta}$, then $\epsilon^{\prime}=1-\left(f_{0}M^{\gamma}+mm_{0}M^{\zeta}\right)/M$. Muscle mass has been shown to be roughly proportional to body mass \cite{} in mammals and thus  $\epsilon^{\prime}$ is effectively $\epsilon$ minus a constant. Thus
\begin{equation}
t_{d}=-\frac{E_{m}\log\left(\epsilon^{\prime}\right)}{B_{m}}
\end{equation}
and $\mu=1/t_{d}$.

It should be noted that we have thus far used mammals an example to describe the size-based relationships for growth, starvation, and death. However, our presentation is general, and other functional forms for $\epsilon$, for example, could be determined for other classes of organisms. Considering bacteria, we might expect that starvation or death is defined by the complete digestion of proteins, and in Table \ref{parameter-values} we provide all parameter values for bacteria which we later use as a comparison in our analysis.

Completing our allometric analysis, for the rate of becoming full, we again use the balance given in Equation \ref{balance} to find the timescale to return to the mature mass from a given reduced starvation mass. The general solution to Equation \ref{balance} is given by
\begin{equation}
m\left(t\right)=\left[1-\left(1-\frac{b}{a}m_{0}^{1-\alpha}\right)e^{-b\left(1-\alpha\right)t}\right]^{1/\left(1-\alpha\right)}\left(\frac{a}{b}\right)^{1/\left(1-\alpha\right)}
\end{equation}
with $a=B_{0}/E_{m}$ and $b=B_{m}/E_{m}$. We are then interested in the timescale, $t_{f}=t_{2}-t_{1}$, which is the time it takes to go from $m\left(t_{1}\right)=\epsilon M$ to $m\left(t_{2}\right)=M$, which has the final form of 
\begin{equation}
t_{f}=\frac{\log \left(1-\left(M \left(\frac{a}{b}\right)^{\frac{1}{\alpha -1}}\right)^{1-\alpha }\right)-\log \left(1-\left(M \epsilon  \left(\frac{a}{b}\right)^{\frac{1}{\alpha
   -1}}\right)^{1-\alpha }\right)}{(\alpha -1) b}.
\end{equation}

\begin{table}[h]
\caption{Parameter Values For Various Classes of Organisms}
\label{parameter-values}
    \begin{center}
    \small
     \begin{tabular}{| p{1.2cm}| p{3.2cm} | p{2.6cm} | p{3.2cm} | }
     \hline
     & {\bf Mammals} & {\bf Unicellular Eukaryotes} & {\bf Bacteria} \\
     \hline
   $\eta$ & $3/4$ & & $1.70$ \\ 
   $E_{m}$ & $10695$ (J gram$^{-1}$) & & $10695$ (J gram$^{-1}$) \\ 
   $E_{m}^{\prime}$ & $\approx E_{m}$ & & $\approx E_{m}$ \\ 
   $B_{0}$ & $0.019$ (W gram$^{-\alpha}$) & & $1.96\times10^{17}$ \\
   $B_{m}$ & $0.025$ (W gram$^{-1}$)   & & $0.025$ (W gram$^{-1}$)\\
   $a$ & $1.78\times10^{-6}$ & & $1.83\times10^{13}$ \\ 
   $b$ & $2.29\times10^{-6}$ & & $2.29\times10^{-6}$ \\  
   $\eta-1$ & $-0.21$ & & $0.73$ \\ 
   $\lambda_{0}$ & $3.39\times10^{-7}$ (s$^{-1}$ gram$^{1-\eta}$) & & $56493$ \\ 
   $\gamma$ & $1.19$ & & $0.68$ \\ 
   $f_{0}$ & $0.02$ & & $1.30\times10^{-5}$ \\ 
   $\zeta$ & $1.01$ & & \\ 
   $mm_{0}$ & $0.32$ & & \\

   \hline
    \end{tabular}
    \end{center}
   \end{table}

%Population growth requires that individuals

@@

%Hopf bifurcation and oscillations sigma vs. rho
The 2-stage consumer resource model exhibits two qualitatively different behavioral regimes.
Because portions of the consumer population exist in either full or hungry states simultaneously, the internal fixed point can either be a stable equilibrium, or exhibit sustained oscillatory behavior, depending on the rates of starvation $\sigma$ and recovery $\rho$.
The transition from stable non-oscillatory dynamics to oscillatory dynamics occurs at the Hopf bifurcation condition where two complex conjugate eigenvalues cross the imaginary axis and attain positive real parts.
Although there is an analytical solution to the Hopf bifurcation condition, it cannot be written efficiently.

When the starvation rate is low, oscillatory dynamics are more likely to occur for a given value of the recovery rate.
This can be understood intuitively: for low starvation rates, resources are depressed by an infusion of full consumers, which subsequently starve thereby allowing the resource to recover and continuing the cycle.
When the starvation rate is high, the response of consumer growth to resource abundance is muted, such that oscillations tend to decay over time.
Thus, higher starvation rates $\sigma$ desensitizes changes in the consumer population to changes in resource density, and lower rates of recovery $\rho$ amplifies this effect.

@@

%The reproductive tradeoff as a strategy
%
\newpage

@@

 \begin{figure}[h]
 	\centering
 	\includegraphics[width=0.5\textwidth]{fig_HopfPlotb.pdf}
 	\caption{
 	}
 	\label{Hopfb}
 \end{figure}

  \begin{figure}[h]
 	\centering
 	\includegraphics[width=0.5\textwidth]{fig_TS.pdf}
 	\caption{
 	}
 	\label{Hopfb}
 \end{figure}

%
%
% \begin{figure}[h]
% 	\centering
% 	\includegraphics[width=0.5\textwidth]{fig_resvuln.pdf}
% 	\caption{
% 	Probability that the resource value is less than threshold = 0.05
% 	}
% 	\label{SN}
% \end{figure}
%
%
%
% \begin{figure}[h]
% 	\centering
% 	\includegraphics[width=0.5\textwidth]{fig_Competition.pdf}
% 	\caption{
% 	Hopf bifurcation
% 	}
% 	\label{SN}
% \end{figure}

\newpage

\section*{References}
